--- 
---

# Project Scoping 

## Overview

By the end of this module, you will: 

- X spatial data visualizations in public health
- X your stakeholders’ needs
- Select the best spatial visualization type for your project 

Think of spatial data visualizations as your bridge to meaningful conversations in public health. Nevertheless, crafting insightful maps, dashboards, or other spatial representations can be challenging, due to both technical and methodological hurdles. The path to creating an effective spatial data visualization does not commence with software usage; rather, it initiates with stakeholder meetings, annotations, and sketches. Therefore, the primary objective of this module is to encourage you to invest time in thinking about your potential visualization and considering the individuals who will reap its benefits. 

Annotations, sketches, and meeting summaries can be handwritten in a notebook or digitally generated on a computer. We recommend creating a journal for this toolkit to annotate and store your notes. Your toolkit journal will assist you in organizing your thoughts, creating new content, making sketches, saving code, writing notes from meetings, and reflecting on your mapping process. Indeed, the production of relevant visualizations related to SDOH is a lengthy and complex process that requires creativity, organization, and technological skills. Your journal can be an invaluable companion on your journey.

### Required Tools {-}

For this module, you need your toolkit journal: 

- Notebook and writing utensil or
- Note-taking or sketching app (if you prefer digital annotations) 

Some of our team members prefer drawing on post-it notes, tablets, digital apps, or wall-sized pieces of paper. Take your pick and go with it!

## What's the Point?

The goal of this project will be to develop some place-based data viz/app/decision support that will integrate social determinants of health and an equity framework.

But, what will you, do, specifically — and why? Let's run through some core questions.

**Where** is your application? Is it within a city, neighborhood, or across an entire country? 

- Greater *spatial coverage* does not necessarily mean that your project will get more complex. It may be easier to grab data for all communities within a city from a data portal, for example, than extract data for a single neighborhood. Having more areas available for comparison can also be useful.

- Greater *spatial resolution* does tend to make projects more complicated, on the other hand. You want to explore state policies for 50 states? No problem. Want to grab census tract data for the entire country (all 77,000)? This will limit your visualization options, as some software approaches will be able to handle it with ease, whereas others not all. 

**Who** is your application *about* — a specific population, defined by their residence and/or some social, economic, or other characteristic? And, **Who** is your application *for*? Who will actually be using your application? How and why are they different?

- Data may not be available for all places and populations. For example, COVID-19 data by race and ethnicity is still not disaggregated below the state-level for all counties in the U.S (!). You may need to start with your ideal, and then work towards what's actually doable.

- As this toolkit will encourage more than once, it is helpful to work with persons who will actually use the application, to increase its meaning and utility. At the same time, it is also helpful to work with persons who the data is about, as lived experiences will provide invaluable understanding, depth, and insight into the project of interest. 

**When** is your application — are you focused on the most recent datasets you can find, or are you interested in data over many time periods? And, how will the application itself be changed or updated over time? 

- Consider the sustainability and long-term maintenance needs of your project. Where will it be in five years? Would it be okay for the application to exist as a time capsule for a project completed, or do you prefer it to have data updated regularly?

**What** actions should/could result from your application? Are they reasonable, feasible, and time-bound? 

- Do you intend for your interactive asset map of cooling centers to be used by community members during a heat event to find refuge? Are you trying to inspire action from policymakers after exploring your data dashboard of modeled map findings and statistics on health disparities? 

- It is common to start with too ambitious a goal, that will slowly get refined over the project's course of implementation. Try to be realistic and honest when assessing what your final project could do.

**Why** is it important? **Why** should people care?

- This is often the most important question, and the one least formulated in early stages of the process. Something may be interesting and meaningful to you, but may not be important to others. Or it may be very obvious to you and your peers, but the thought hasn't occurred to others in a different discipline or domain. Be specific and intentional in setting this objective, and do the research. 

Put these together to generate an overview goal statement for your project. Include the primary question you’re trying to answer. This may change throughout the process, but should be changed to be refined further, rather than expanded in scope.  

Think of the start of your project as setting out on a road trip. First, you need to map out where you want to go. What do you hope to discover along the way? Your visualization techniques are like choosing the right vehicle for the journey—each has its strengths for different terrains. Your goals will evolve as you travel, finding new paths to explore. But it all starts with a clear destination in mind.

## Refining Goals for SDOH 

Taking your project a step further, dig deeper into how your application can address SDOH directly. Which common approach or technique is the most pertinent to your goals? Consider the following ideas:

 <details markdown='1'><summary>**Identifying Health Disparities**</summary> You can employ spatial visualizations to identify areas with disproportionate health burdens and disparities. For example, by mapping health outcomes and demographic data from the census, it could become evident which communities or populations experience higher rates of diseases or poor health, shedding light on potential health inequities. </br></br></details>
 
  <details markdown='1'><summary>**Targeting Interventions**</summary>  Another option is performing spatial analyses to target interventions and resources to areas with the greatest need. By creating asset maps, you could aid public health advocates to direct their efforts, and allocate resources to address health disparities and promote equity. </br></br></details>
 
  <details markdown='1'><summary>**Assessing Access to Healthcare**</summary> You can produce maps to evaluate the accessibility of healthcare services across different regions. These cartographies identify areas with limited access to medical facilities or services, which is crucial in understanding barriers to healthcare for marginalized populations and addressing health disparities. </br></br></details>
  
   <details markdown='1'><summary>**Modeling SDOH at Different Scales**</summary> You can produce geostatistical analyses to model how social determinants of health (SODH) affect communities at different scales. By studying and mapping these factors, you may better understand how social conditions influence health outcomes and equity. </br></br></details>
   
   <details markdown='1'><summary>**Environmental Justice**</summary> You could create a dashboard to assess environmental health risks and exposures. By combining graphs, maps, and tables you may identify areas with environmental hazards that disproportionately affect specific communities, contributing to health disparities. </br></br></details>
    
  <details markdown='1'><summary>**Health Planning and Policy Interventions**</summary> Your maps can inform stakeholders in their efforts to create effective policies. By deploying various spatial visualizations, you can support interventions targeting specific health challenges in various communities, ultimately promoting health equity.</br></br></details>
  
  <details markdown='1'><summary>**Community Engagement**</summary> Story maps or collaborative mapping projects may facilitate community engagement by providing social experience insights. These representations allow community members to participate in decision-making, voice their concerns, and collaborate with public health officials to design interventions that address their unique needs and concerns. </br></br></details>

  <details markdown='1'><summary>**Monitoring and Evaluation**</summary> Spatio-temporal dashboards are effective tools to monitor the effectiveness of interventions over time. By comparing health outcomes before and after implementing interventions, public health officials can assess whether disparities are decreasing and health equity is improving.</br></br></details>

</br>
Take a moment to consider which goal(s) you have for your project. We recommend starting simple, with no more than one or two goals to start with. Your initial project should focus on accomplishing your primary goal to the greatest of your abilities. Over time, more experience, resources, and support will enable you and your team to expand further.
</br>


```{r table-simple, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
require(pander)
panderOptions('table.split.table', Inf)
my.data <- "  Goal        | Description
  Identifying Health Disparities      | Map health outcomes & demographics to identify disparities.
  Targeting Interventions     | Use spatial analysis to direct resources effectively.
  Assess Access to Healthcare | Map service accessibility to pinpoint healthcare gaps.
  Modeling SDOH at Different Scales | Analyze SDOH imapct across scales for insight into health influences.
  Environmental Justice | Create dashboards to monitor enviornmental health risks.
  Health Planning & Policy Interventions | Inform policy with visualizations to address health challenges.
  Community Engagement | Use story maps to involve communities in decision-making.
  Monitoring and Evaluation | Track interventions over time with spatio-temporal dashboards."
set.caption("Common Goals for SDOH & Place Applications")
df <- read.delim(textConnection(my.data),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
names(df) <- unname(as.list(df[1,])) # put headers on
df <- df[-1,] # remove first row
row.names(df)<-NULL
pander(df, style = 'rmarkdown')
```
</br>

### Integrate Equity Directly

Think back to the equity framework introduced in the first module. While each component is critical to any application built to understand and communicate SDOH, the [**Content of Equity**](https://maps.healthyregions.org/sdhoplace-toolkit/01-intro.html#content-of-equity) component can be a useful consideration when building into your goals. Consider how your application may address equity by focusing on one or more of these aspects. Some examples include:

- Integrating community stakeholders directly to the decision-making processes, planning, and implementation of your web application. *(Procedural Equity)*
- Developing an asset map of local health and social services to maximize networking and connection in your city. *(Distributive Equity)*
- Curating a story map that invites users to "walk" through a neighborhood over time as it becomes segregated from highway construction. *(Contextual Equity)*
 
### Activity
Take a look at these three real-world projects. In your own words, sketch out their goals. Do they align with any of the SDOH application goals we've mapped above? What content of equity (procedural, distributive, or contextual) do they address?

- [Data and Stats | Tulsa Health Department](https://www.tulsa-health.org/community-health/data-stats/north-tulsa-medical-provider-and-pharmacy-asset-map)

- [United States of Health Dashboard | McKinsey & Company](https://www.mckinsey.com/industries/public-sector/our-insights/us-public-health-dashboard)

- [Heart Disease Death Rates Aged 35 and older, Total Population | cdc.gov](https://www.cdc.gov/dhdsp/maps/national_maps/hd_all.htm)

Next, update your own goal statement based on ideas in this section. What are your SDOH and Equity goals?

### Project Tip 

Module 1 is just the beginning of a process. To keep progressing, it's essential to always look for inspiration and stakeholders. Take some time to explore various web maps, dashboards, atlases, magazines, and academic articles. Then, list the ones you find appealing in your journal. Even if they are spatial visualizations that have nothing to do with health, try to imagine how you can use their design to study SDOH.

## Defining Stakeholders 
Once you've got your project's goals on the horizon, it's time to look at who's journeying with you. Collaboration and stakeholder identification are vital components of any project that aims to achieve equity. Whether it's a non-governmental organization, a research group within a university, or a government agency, stakeholders bring diverse perspectives and expertise to the table. Similarly, communities, activists, and advocates can also play a significant role in your project's success. 

<blockquote>Early identification of stakeholders is key to building trust and solidarity, which is a time-consuming but necessary process. </blockquote>

In Module 2, we will delve deeper into employing User-Centered Design to engage stakeholders effectively. The aim will be to ensure that every step of the project, from conception to implementation, is done in collaboration with those who have a vested interest in the outcomes, guaranteeing that the resulting initiatives are equitable, impactful, and sustainable. For now, let's assess your team.


### Assess Your Team

Take an inventory of who makes up your team, and define each person's role. If you are embedded within a larger organization or corporation, there may be expansive stakeholders involved, with multiple stages of review, implementation, and iteration. If you're in a smaller group or you're on your own, you may be taking on several roles at the same time. Your core team may include community members directly, or may have a different leadership structure. Here are some critical roles to consider:

```{r table-simple2, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
require(pander)
panderOptions('table.split.table', Inf)
my.data <- "  Responsibilities        | Description
  Set Strategy, Priorities, & Standards     | Core Team, Leadership
  Allocate Resources (Time and $)     | Leadership
  Manage Processes | Core Team, Project Manager 
  Design, Research Use Cases | Core Team, Designer
  Develop, Deploy, & Monitor Project | Core Team, Engineer
  Use the application in way that was expected | Expected Users
  Use the application in way that was uexpected  | Unexpected Users
"
df <- read.delim(textConnection(my.data),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
names(df) <- unname(as.list(df[1,])) # put headers on
df <- df[-1,] # remove first row
row.names(df)<-NULL
pander(df, style = 'rmarkdown')
```

There are two additional stakeholders every application developer should consider -- the Champion, and the Curmudgeon. Yes, these are actual roles that have been used in *Health Informatics* teaching for some time... Considering them will be essential for your success:

- **The Champion** - They love your idea, sing its praises, and are ready to share your work with everyone they know. They may not get the "tech stuff" completely, and won't have time to learn it if they wanted to, but are your biggest cheerleader. They may help you get funding.
- **The Curmudgeon** - They prefer to do things the way they have always been doing it. It's generally something they do themselves, and probably manually. They are not interested in your work, and may challenge it with a sigh, eye-roll, or outright lament. May prefer technology the way it "used" to be.

We caution that you not shift your work too much in either direction to please/antagonize either of these stakeholders, but rather understand their role in the process at this stage. 

### Motivations
If you find yourself to be a "curmudgeon" at the idea of working with others to support your journey, we also encourage you to take a moment to consider all the benefits of engaging more stakeholders.

Here’s how a wider stakeholder alliance can shape your journey:

- **Understanding Diverse Perspectives:** Stakeholders come from various backgrounds and sectors, each bringing unique insights that can challenge and enrich your project's approach to health. A government official might understand policy implications, while a local activist might provide a grassroots perspective on community health challenges.
- **Data Gathering and Validation:** Collaborators are crucial in both collecting and validating SDOH data. They ensure that the information reflects real-world conditions and is relevant to the communities affected. Researchers can provide rigorous methodologies, while community members can offer qualitative insights that ground the data in lived experiences.
- **Expanding the Network:** Engaging with stakeholders allows for the expansion of your network. Each stakeholder may introduce you to other relevant parties, broadening the reach and potential impact of your project. This can lead to discovering untapped resources or finding new avenues for support and advocacy.
Exploring New Contexts: Stakeholders can act as guides through unfamiliar territories, both metaphorically in the project domain and literally in the community spaces. They help navigate the cultural, social, and political landscapes that shape health outcomes.
- **Innovative Co-Design:** Utilizing User-Centered Design principles, stakeholders become co-creators in developing spatial visualizations that represent SDOH. Their input ensures that the end products are not only technically sound but also culturally sensitive and user-friendly.
- **Forging Partnerships:** Collaboration can lead to long-term partnerships that extend beyond the life of a single project. These relationships can build a foundation for future initiatives, creating a sustainable impact on public health.
Revising Project Goals:** Stakeholders often provide new insights that can lead to a reframing of project goals. What begins as a narrow focus on a particular health issue might evolve into a more holistic approach that considers a wider range of social factors.
- **Ensuring Equity:** By involving a diverse group of stakeholders, the project is more likely to address equity in a meaningful way. Equity is not just about equal access, it's also about designing interventions that acknowledge and address power imbalances and historical contexts that contribute to health disparities.

### Finding a Community of Practice
Even if you think you're getting into this solo or with a smaller team, you may be surprised at the number of people available to connect with you as invaluable colleagues, future friends, and mentors in your network. You've found this project -- that already means you're a part of a wider community of practice. Join our [**LinkedIn**](#) and [**GitHub**](#) pages to introduce yourself and expand your network.

Not sure where to start looking for mentors? We love this [**Mentor Map**](https://www.insidehighered.com/sites/default/files/files/Mentoring%20Map%5B1%5D(1).pdf) exercise by the *National Center for Faculty Development & Diversity* -- not just for academic folks, but anyone who has a beating heart and is working on a project. Try out the exercise yourself to identify persons in your network who you may be reach out to connect for feedback, accountability, intellectual discussions, and emotional support as you work through the project. 

### Activity

Review these two projects and identify their stakeholders. Explain what is their role in each project.

- [Urban Youths Go 3000 Miles: Engaging and Supporting Young Residents to Conduct Neighborhood Asset Mapping | AJPH](https://ajph.aphapublications.org/doi/full/10.2105/AJPH.2011.300351)
- [Food assets for whom? Community perspectives on food asset mapping in Canada](https://www.tandfonline.com/doi/full/10.1080/17549175.2021.1918750)


## Project Assessment

You now know what you want to do, and who you want to do it with. You've gathered your team and had a few good discussions. If you're on your own, you've made a list of mentors and additional resources to connect with so you can ensure feedback throughout the process.

Let's finish scoping this project.

### Assess Your Strengths

Each project will require **domain expertise** and **technical capacity**. Domain expertise means knowledge ... Technical capacity will refer to ...

Which is your strength, and which will you need to skill up to further support for project implementation? 

For example: 

- If your project is about building an asset map to support populations experiencing food insecurity, you may be a food bank volunteer who has worked directly with populations impacted, live in the community, and also would be using the application with collaboration from the regional food pantry network. In this case, you have extensive domain expertise and knowledge that will be more enriched than most web app developers, assuming you are not a developer for your day job. For this project, your goal will be to scale up your technical capacity within the constraints of your setting. 

- On the other hand, perhaps you are a graduate student who has been coding statistical models for your work, and are interested in adding more social impact to your work. You’re used to learning new programming languages for new projects, and have a high technical capacity, but are not embedded within the community you want to build an asset map for. For this project, your goal will be to scale up your understanding of equity based frameworks, learning more about community experiences, as well as researching the topic further to build your domain expertise. 

After assessing your strengths, go back to your list of stakeholders and update as needed. Note that as your expand your stakeholder team, your project may update according to shared goals and visions of team members. As a result, your team will work through the assessment process more than once. This iterative process is expected.

### Assess Project Needs
Project Needs will include data, computer services, time, and attitude. To assess this, you will need to survey your motivation and paint points, as well as technical, financial, people, and time resources. Consider sustainability at this stage… 

### Establish Deliverables

Technial requirements

++ application action goals:
- Access to information (ex. Drop-down selection, click for pop-up window, scrollytelling)
- User-prompted input (ex. Slider, number entry)

++ SMART goals/similar? 

### Establish Process



++ data preparation, analysis, application development (recommend that just 1 is intensive)

Project Manage

- Generate a timeline, and break out tasks
- How will you manage & document your process?
- Communications plan between team members & stakeholders?

Agile Processes




### Module 1 Activity: {-}

Think of a problem you want to solve. Write down your purpose and objectives by clearly stating why you want to solve this problem and what you hope to achieve. Who is your target audience? 

Next, make a list of your potential stakeholders. How would you contact your stakeholders and why would they be interested in collaborating with you? Would you be part of an ongoing project or propose a new one?

The format for this exercise is flexible and exploratory. You can opt to create lists of objectives, draw diagrams, write a structured plan, or engage in a free-writing exercise.

The primary aim of this task is to develop a foundation to begin exploring spatial data for your project.It's important to note that your final prototype might diverge significantly from what you record in this entry, and that's perfectly acceptable. Remember, the process of using SDOH spatial data for visualizations is a dynamic process.




#### Summary of Module 1 {-}

In Module 1, the focus is on goal setting for a spatial visualization project related to social determinants of health (SDOH) in public health. Here's a summary of the key points:

Objectives (1.1):
- Clarifying project goals is the initial step.
- The choice of visualization techniques is closely tied to project objectives.
- General goals include identifying health disparities, targeting interventions, assessing access to healthcare, modeling SDOH at different scales, evaluating environmental health, informing health planning and policy, engaging the community, and monitoring and evaluating interventions.

Stakeholders (1.2):

- Collaboration and stakeholder identification are essential for achieving equity.
- Stakeholders can be organizations, researchers, government agencies, communities, activists, and advocates.
- Early identification of stakeholders builds trust and solidarity.

Spatial Data (1.3):
- Social determinants of health encompass conditions in which people are born, grow, live, work, and age.
- Spatial data is crucial for understanding how communities are affected by their environments.
- Three types of spatial data are primary (collected personally), secondary (from databases and census records), and tertiary (previously used in other maps).

Data Visualizations (1.4):
- Four types of spatial visualizations are discussed: asset maps, data dashboards, story maps, and thematic maps.
- Asset maps focus on community strengths and resources.
- Data dashboards present health-related data in a user-friendly format.
- Story maps combine maps, text, images, and multimedia for storytelling.
- Thematic maps use visual symbols to represent specific health-related data or themes within a geographic area.

The module also encourages exploring external resources and encourages participants to think about their own project's purpose, objectives, target audience, and potential stakeholders. It emphasizes that Module 1 is the beginning of a dynamic process in the journey of using spatial data for SDOH visualizations in public health.
